---
title: "MLB Season 23 xwOBA Modeling Attempts"
output:
  html_document:
    df_print: paged
classoption: landscape
---

```{r setup, include=FALSE, echo = FALSE}
library(tidyverse)
library(knitr)
library(caret) # for KNN
# library(mgcv) # for GAM (maybe not)
library(dplyr) # for df stuff
library(baseballr) # for statcast queries
library(VGAM) # for GAM
library(ggplot2)
```

```{r, include = FALSE}
# Read the Data
mlb_reg_21 <- read.csv(file = 'mlb_reg_21.csv')
mlb_reg_22 <- read.csv(file = 'mlb_reg_22.csv')
mlb_reg_23 <- read.csv(file = 'mlb_reg_23.csv')

# Download sprint speed - two birds with one stone get batter names + speed for model
speed_21 <- statcast_leaderboards(leaderboard = "sprint_speed", year = 2021)
speed_21 <- speed_21 %>%
  rename(batter_name = `last_name, first_name`,
         batter = player_id) %>%
  select(batter_name, batter, sprint_speed)

mlb_reg_21$game_date <- as.Date(mlb_reg_21$game_date)
mlb_reg_21 <- merge(mlb_reg_21, speed_21)

speed_22 <- statcast_leaderboards(leaderboard = "sprint_speed", year = 2022)
speed_22 <- speed_22 %>%
  rename(batter_name = `last_name, first_name`,
         batter = player_id) %>%
  select(batter_name, batter, sprint_speed)

mlb_reg_22$game_date <- as.Date(mlb_reg_22$game_date)
mlb_reg_22 <- merge(mlb_reg_22, speed_22)

speed_23 <- statcast_leaderboards(leaderboard = "sprint_speed", year = 2023)
speed_23 <- speed_23 %>% 
  rename(batter_name = `last_name, first_name`,
         batter = player_id) %>% 
  select(batter_name, batter, sprint_speed)

mlb_reg_23$game_date <- as.Date(mlb_reg_23$game_date)
mlb_reg_23 <- merge(mlb_reg_23, speed_23)
```

```{r}
mlb_reg_2122 <- rbind(mlb_reg_21, mlb_reg_22)
```


I am using the following resource to base my calculations on: https://technology.mlblogs.com/an-introduction-to-expected-weighted-on-base-average-xwoba-29d6070ba52b

# Overview

xwOBA is a derived statistic that comes from the following formula:

$$xwOBA = (xwOBAcon + wBB \times (BB-IBB) + wHBP \times HBP)/(AB + BB - IBB + SF + HBP)$$

Here, $xwOBAcon$ is itself a modeled statistic for $xwOBA$ on contact only, and $wSTAT$ indicates the $wOBA$ weight.

# xwOBAcon

We should try a model based on the following variables: exit velocity (EV), launch angle (LA), and sprint speed.

"Our newest model is a combination of k-NN and GAMs. We use GAMs to model most weakly hit balls, shallow infield pop-ups, and grounders, the idea being that these are balls where speed matters. We apply our existing k-NN model to the remaining liners and fly balls (where the batter’s speed has far less impact on the outcome), only using EV and LA to estimate wOBA."

```{r}
rm(mlb_reg_21, mlb_reg_22, speed_21, speed_22, speed_23)
```


## First thing is first: Data filtering and Exploratory Analysis

TBH may switch up sources here: https://medium.com/@thomasjamesnestico/modelling-xwoba-with-knn-9b004e93861a

```{r}
# Filter just contact events
# Do we include shit like fielders choice etc
mlb_con_2122 <- mlb_reg_2122 %>% select("events", "launch_speed", "launch_angle", "sprint_speed", "hit_distance_sc", "bb_type", "batter_name", "game_date") %>% filter(events %in% c("field_out", "field_error", "double_play", "grounded_into_double_play", "single", "double", "triple", "home_run", "sac_fly", "sac_fly_double_play"))
# Turn all random ish into field out
mlb_con_2122$events <- ifelse(mlb_con_2122$events %in% c("single", "double", "triple", "home_run"), mlb_con_2122$events, "field_out")
# mlb_con$events <- ifelse(mlb_con$events == "field_out", 0, mlb_con$events)
# mlb_con$events <- ifelse(mlb_con$events == "single", 1, mlb_con$events)
# mlb_con$events <- ifelse(mlb_con$events == "double", 2, mlb_con$events)
# mlb_con$events <- ifelse(mlb_con$events == "triple", 3, mlb_con$events)
# mlb_con$events <- ifelse(mlb_con$events == "home_run", 4, mlb_con$events)
mlb_con_2122$events <- as.factor(mlb_con_2122$events)

mlb_con_2122 <- mlb_con_2122 %>% filter(!is.na(launch_speed)) %>% filter(!is.na(launch_angle))

mlb_con_23 <- mlb_reg_23 %>% select("events", "launch_speed", "launch_angle", "sprint_speed", "hit_distance_sc", "bb_type", "batter_name", "game_date") %>% filter(events %in% c("field_out", "field_error", "double_play", "grounded_into_double_play", "single", "double", "triple", "home_run", "sac_fly", "sac_fly_double_play"))
# Turn all random ish into field out
mlb_con_23$events <- ifelse(mlb_con_23$events %in% c("single", "double", "triple", "home_run"), mlb_con_23$events, "field_out")
# mlb_con$events <- ifelse(mlb_con$events == "field_out", 0, mlb_con$events)
# mlb_con$events <- ifelse(mlb_con$events == "single", 1, mlb_con$events)
# mlb_con$events <- ifelse(mlb_con$events == "double", 2, mlb_con$events)
# mlb_con$events <- ifelse(mlb_con$events == "triple", 3, mlb_con$events)
# mlb_con$events <- ifelse(mlb_con$events == "home_run", 4, mlb_con$events)
mlb_con_23$events <- as.factor(mlb_con_23$events)

mlb_con_23 <- mlb_con_23 %>% filter(!is.na(launch_speed)) %>% filter(!is.na(launch_angle))
```

```{r}
# Exploratory Plots
# ggplot(data = mlb_con) +
#   geom_point(aes(x = launch_speed, y = launch_angle, color = events))
```

```{r}
# Need to now separate out Sprint Speed influenced BBE, and non sprint speed (GAM and KNN)
mlb_soft_con_2122 <- mlb_con_2122 %>% filter((launch_speed < 75)& (launch_angle < 60)&(hit_distance_sc < 120))
mlb_hard_con_2122 <- mlb_con_2122 %>% anti_join(mlb_soft_con_2122)

mlb_soft_con_23 <- mlb_con_23 %>% filter((launch_speed < 75)& (launch_angle < 60)&(hit_distance_sc < 120))
mlb_hard_con_23 <- mlb_con_23 %>% anti_join(mlb_soft_con_23)
```

## Split into testing/training

```{r}
# # Split into test and train as well as some knn setup
# 21-22 train, 23 test

# Run algorithms using 10-fold cross validation
trainControl <- trainControl(method="repeatedcv", number=10, repeats=3)
metric <- "Accuracy"
```

## KNN Model for hard hit (relatively) BBE

```{r}
set.seed(7)
fit.knn <- train(events~launch_speed+launch_angle, data=mlb_hard_con_2122, method="knn",
                 metric=metric ,trControl=trainControl)
knn.k1 <- fit.knn$bestTune # keep this Initial k for testing with knn() function in next section
print(fit.knn)
```

```{r}
plot(fit.knn)
```

```{r}
set.seed(7)
prediction <- predict(fit.knn, newdata = mlb_hard_con_23)
cf <- confusionMatrix(prediction, as.factor(mlb_hard_con_23$events))
print(cf)
```

## From KNN to useful stuff:

Must turn these predicted class probs into a digestible number, our $xwOBAcon$. To do so, lets take the predicted class probabilites, and find the sum

```{r}
predicted_probs_knn <- predict(fit.knn, newdata = mlb_hard_con_23, type = "prob")
```


```{r}
single_factor <- 0.855
double_factor <- 1.248
triple_factor <- 1.575
hr_factor <- 2.014

predicted_probs_knn$field_out <- 0
predicted_probs_knn$single <- predicted_probs_knn$single * single_factor
predicted_probs_knn$double <- predicted_probs_knn$double * double_factor
predicted_probs_knn$triple <- predicted_probs_knn$triple * triple_factor
predicted_probs_knn$home_run <- predicted_probs_knn$home_run * hr_factor

xwOBAcon_vec_knn <- rowSums(predicted_probs_knn)
```

```{r}
# Add to original df for grouping
mlb_hard_con_23$xwOBAcon_knn <- xwOBAcon_vec_knn
xwOBAcon_player_knn <- mlb_hard_con_23 %>% group_by(batter_name) %>% dplyr::summarize(xwOBAcon_sum_knn = sum(xwOBAcon_knn))
```

## Now repeat the process but for the GAM with sprint speed

```{r}
# Linear sprint term, cubic spline LA and EV ? is this justified
# Fit a multinomial GAM
gam_multinomial <- vglm(events ~ sprint_speed + s(launch_angle, bs = "cs") + s(launch_speed, bs = "cs"), family = multinomial(), data = mlb_soft_con_2122, trace = TRUE)

# Check the summary
summary(gam_multinomial)
```

```{r}
predicted_probs_gam <- predict(gam_multinomial, newdata = mlb_soft_con_23, type = "response")
home_run <- 1 - rowSums(predicted_probs_gam)
predicted_probs_gam <- as.data.frame(cbind(predicted_probs_gam, home_run))

predicted_probs_gam$field_out <- 0
predicted_probs_gam$single <- predicted_probs_gam$single * single_factor
predicted_probs_gam$double <- predicted_probs_gam$double * double_factor
predicted_probs_gam$triple <- predicted_probs_gam$triple * triple_factor
predicted_probs_gam$home_run <- predicted_probs_gam$home_run * hr_factor

xwOBAcon_vec_gam <- rowSums(predicted_probs_gam)
```

```{r}
# Add to original df for grouping
mlb_soft_con_23$xwOBAcon_gam <- xwOBAcon_vec_gam
xwOBAcon_player_gam <- mlb_soft_con_23 %>% group_by(batter_name) %>% dplyr::summarize(xwOBAcon_sum_gam = sum(xwOBAcon_gam))
```

## More data preparation: Find number of AB's, BB's, etc per player

```{r}
# player_summary <- mlb_reg_23 %>% 
#   group_by(batter_name) %>% 
#   dplyr::summarize(HBP = sum(events == "hit_by_pitch", na.rm = TRUE), 
#                    #uBB = sum(events %in% c(), na.rm = TRUE),
#                    #IBB = sum(events %in% c(), na.rm = TRUE),
#                    SF = sum(events %in% c("sac_fly", "sac_bunt"), na.rm = TRUE),
#                    BB = sum(events == "walk"),
#                    AB = sum(events %in% c("double", "double_play", "field_error",
#                                           "field_out", "fielders_choice",
#                                           "fielders_choice_out", "force_out",
#                                           "grounded_into_double_play", "home_run",
#                                           "single","strikeout", "strikeout_double_play",
#                                           "triple"), na.rm = TRUE))
player_summary <- read.csv(file="player_summary_23.csv")
player_summary <- player_summary %>% 
  rename(batter_name = last_name..first_name,
         batter = player_id)
```

```{r}
final_df <- merge(xwOBAcon_player_knn, player_summary, on='batter_name')
final_df <- merge(xwOBAcon_player_gam, final_df, on='batter_name')
final_df$xwOBA <- (0.697*(final_df$walk - final_df$b_intent_walk) + 0.727*final_df$b_hit_by_pitch + final_df$xwOBAcon_sum_knn + final_df$xwOBAcon_sum_gam)/(final_df$ab + final_df$b_sac_fly + final_df$walk - final_df$b_intent_walk + final_df$b_hit_by_pitch)
final_df <- final_df %>% filter(ab > 100)
```

```{r}
final_df
```


```{r}
wOBA <- statcast_leaderboards(leaderboard = "expected_statistics", year = 2023, min_pa = 1)
wOBA <- wOBA %>% 
  rename(batter_name = `last_name, first_name`,
         batter = player_id) %>% 
  select(batter_name, batter, woba, est_woba)
```

```{r}
overall <- merge(final_df, wOBA, on='batter_name')

# vs regular woba
ggplot() +
  geom_point(data = overall, aes(x = woba, y = xwOBA)) +
  geom_segment(aes(x = 0.1, y = 0.1, xend = 0.5, yend = 0.5))

# vs statcast xwoba
ggplot() +
  geom_point(data = overall, aes(x = est_woba, y = xwOBA)) +
  geom_segment(aes(x = 0.1, y = 0.1, xend = 0.5, yend = 0.5))
```

```{r}
overall$xwoba_diff <- overall$xwOBA - overall$est_woba

val_mod <- lm(overall$est_woba ~ overall$xwOBA + 0)
summary(val_mod)
```


```{r}
# need a way to manually scrape
# custom_leaderboards <- statcast_leaderboards(leaderboard = "custom_leaderboards", year = 2023, min_pa = 1)
```

```{r}
saveRDS(fit.knn, file="KNN-xwOBA_(hard_hit).rda")
saveRDS(gam_multinomial, file="GAM-xwOBA_(soft_hit).rda")
```






* Thinking emoji * https://www.reddit.com/r/baseball/comments/15veyzp/baseballs_unseen_bias_how_woba_xwoba_xera_and_wrc/
